### 2020-12-08 ch02_scientific_background/scientific_background.tex
Added overview of climate feedbacks

These differences in the climate models also give rise to different intensities
of \emph{climate feedbacks} (or even their presence/absence) in the models. A
climate feedback is a mechanism that either amplifies (\emph{positive
  feedback}) or weakens (\emph{negative feedback}) the effect of an external
forcing. An example of a positive feedback is the water vapor feedback, where
the increased surface temperature (caused by anthropogenic forcing) leads to an
enhanced evaporation of water which enhances the amount of water vapor in the
atmosphere. Since water vapor itself is a powerful \ac{GHG}, this amplifies the
effect of the anthropogenic forcing by further increasing the surface
temperatures \autocite{Cubasch2013}. Further examples and a mathematical
description of feedbacks are given in
\cref{subsec:02:climate_feedbacks}.

@@

As already described in \cref{subsec:02:source_of_uncertainties}, the effects
of an external forcing acting on the climate system can additionally be
amplified or weakened by climate feedbacks. Thus, feedback processes play a
crucial role in the evaluation of climate sensitivity.

\begin{figure}[t]
  \centering
  \includegraphics[width=\LargeFigureWidth{}]{
    ch02_scientific_background/figs/feedbacks_overview.pdf}
  \caption{Climate feedbacks and corresponding time scales. \enquote{+} refers
    to positive feedbacks, which amplify the effect of the external forcing
    (\eg{} the water vapor feedback). \enquote{-} refers to negative feedbacks,
    which weaken the effect of the external forcing (\eg{} the longwave
    radiation feedback). \enquote{+/-} refers to feedbacks which might be
    either positive or negative (\eg{} the cloud feedback). The smaller box
    highlights the large differences in time scales for the various feedbacks.
    Taken from \textcite{Cubasch2013}.}
  \label{fig:02:feedbacks_overview}
\end{figure}

- Overview over feedbacks (Cubasch et al., 2013, Figure 1.2)

- Timescales of feedbacks

- Longwave radiation feedback = Planck feedback

- Mathematical description of feedbacks, geometric series

- Equations for feedbacks + decomposition

@@

\section{Techniques to reduce Uncertainties in Climate Model Projections}
\label{sec:02:techniques}

>>>>

These differences in the climate models also give rise to different intensities
of \emph{climate feedbacks} (or even their presence/absence) in the models. A
climate feedback is a mechanism that either amplifies (\emph{positive
  feedback}) or diminishes (\emph{negative feedback}) the effect of an external
forcing. An example of a strong positive feedback is the water vapor feedback,
in which the increased surface temperature (caused by anthropogenic forcing)
leads to an enhanced evaporation of water which enhances the amount of water
vapor in the atmosphere. Since water vapor itself is a powerful \ac{GHG}, this
amplifies the effect of the anthropogenic forcing by further increasing the
surface temperatures \autocite{Cubasch2013}. Further examples and a
mathematical description of feedbacks are given in
\cref{subsec:02:climate_feedbacks}.

@@

As already described in \cref{subsec:02:source_of_uncertainties}, the effects
of an external forcing acting on the climate system can additionally be
amplified or diminished by climate feedbacks. Thus, feedback processes play a
crucial role in the evaluation of climate sensitivity.
\Cref{fig:02:feedbacks_overview} shows an overview of important feedbacks in
the Earth system with their corresponding time scales on which they operate.

\begin{figure}[t]
  \centering
  \includegraphics[width=\LargeFigureWidth{}]{
    ch02_scientific_background/figs/feedbacks_overview.pdf}
  \caption{Climate feedbacks and corresponding time scales. \enquote{+} refers
    to positive feedbacks, which amplify the effect of the external forcing
    (\eg{} the water vapor feedback). \enquote{-} refers to negative feedbacks,
    which diminish the effect of the external forcing (\eg{} the longwave
    radiation feedback). \enquote{+/-} refers to feedbacks which might be
    either positive or negative (\eg{} the cloud feedback). The smaller box
    highlights the large differences in time scales for the various feedbacks.
    Taken from \textcite{Cubasch2013}.}
  \label{fig:02:feedbacks_overview}
\end{figure}

An example for a positive feedback is the already mentioned \emph{water vapor
  feedback}. Being the primary \ac{GHG} in the Earth's atmosphere, water vapor
is the largest contributor to the natural greenhouse effect. Since its amount
in the atmosphere is mainly controlled by the air temperature and anthropogenic
emissions of water vapor are negligible, the influence of water vapor on the
climate system is described as a feedback mechanism and not as an external
forcing \autocite{Myhre2013}. Basis of this feedback is the enhanced
evaporation of water with increasing air temperatures. Each degree of warming
allows the atmosphere to retain about \pct{7} more water vapor
\autocite{Myhre2013}, which closes the positive feedback loop by further
increasing air temperatures through the greenhouse effect. With a typical
residence time of water vapor in the atmosphere of several days, the water
vapor feedback operates on relatively short time scales. As the largest
positive feedback in the Earth system \autocite{Soden2006}, the water vapor
feedback amplifies any initial forcing (\eg{} caused by anthropogenic \ac{CO2}
emissions) by a typical factor between two and three, rendering water vapor a
fundamental agent of climate change \autocite{Myhre2013}. An example for a
positive feedback that operates on longer time scales (several years) is the
\emph{snow/ice albedo feedback}, in which the surface albedo decreases as
highly reflective ice and snow surfaces melt with global warming, exposing the
darker and more absorbing surfaces below \autocite{Cubasch2013}.

In contrast to positive feedbacks, negative feedbacks diminish the effect of an
external forcing. An example for this is the \emph{blackbody feedback} (also
known as \emph{Planck feedback} or \emph{longwave radiation feedback}), which
is the strongest negative feedback in the Earth system \autocite{Cubasch2013}.
It is based on the thermal electromagnetic radiation that any object with a
non-zero temperature emits (the so-called \emph{blackbody radiation}). Since
the power of this radiation strongly depends on the temperature of the object,
higher surface temperatures of Earth increase the outgoing longwave radiation
flux from the surface which reduces the effect of the external forcing and
cools the planet.

For some domains of the Earth system, feedbacks can be positive and/or
negative, since a variety of different mechanisms is involved. An example for
this is the cloud feedback. Changes in clouds induced by climate change can
cause both longwave (greenhouse warming) and shortwave (reflective cooling)
effects, which both need to be considered for the overall cloud feedback
\autocite{Boucher2013}. Relevant cloud properties that may change as a response
to an external forcing and that may alter the Earth's radiative budget are
cloud cover, cloud optical thickness, cloud altitude and the geographical
distribution of clouds. Examples for robust cloud feedback processes are the
rise of high clouds in a warming climate which traps longwave radiation and
enhances global warming and the reduction in mid- and low-level cloud cover
which diminishes the reflection of incoming solar radiation and also increases
the surface warming \autocite{Boucher2013}. In global climate model ensembles,
the overall cloud feedback shows a large range with positive and negative
values, but tends to be slightly positive on average \autocite{Soden2006,
  Dufresne2008, Vial2013, Zelinka2020}. This large uncertainty in the cloud
feedback is a major reason for uncertainties in the climate sensitivity of
climate models \autocite{Boucher2013, Flato2013}.

Further examples of feedbacks with positive and negative contributions are
biogeochemical feedbacks. Negative contributions come from increased \ac{CO2}
fluxes into the land and ocean carbon reservoirs due to increased
photosynthesis rates and \ac{CO2} dissolution in the sea, which decrease the
atmospheric \ac{CO2} content and diminish global warming. An example for a
positive contribution is the decreased solubility of \ac{CO2} in water in a
warmer climate, which reduces the atmosphere-ocean \ac{CO2} flux and enhances
climate change. More details on this are given in
\cref{subsec:02:carbon_cycle_perturbations}.

\subsection{Mathematical Description of Feedbacks}
\label{subsec:02:mathematics_of_feedbacks}

- Mathematical description of feedbacks

- Equations for feedbacks + decomposition (reference for feedbacks: Planck, Roe
2009)

@@

\section{Techniques to Reduce Uncertainties in Climate Model Projections}
\label{sec:02:techniques}
